---
title: "average-daily"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{average-daily}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MSD)
```

The MSD package is designed to analyze one year of data at a time. For the AGU '23 conference, a graphic was made that showcased the trends of the average rainfall across a set of years. That is, each day's rainfall data across multiple years were averaged, then analyzed. For example, in a data set with precipitation data from 1981 to 1985, every day of the year (from January 1st to December 31st) had its rainfall data averaged (example: the precipitation values of 1-1-81, 1-1-82, 1-1-83, 1-1-84, and 1-1-85 were averaged), then put into a dummy xts to be analyzed. This vignette shows how that graphic was created.

The first step is to extract the relevant data from a SpatRaster.

```{r}
library(MSD)
library(terra)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(xts)
data = system.file("data", "prcp_cropped.tif", package="MSD") #This loads the data included in the package, but you would attach your own
infile = terra::rast(data)

# Make a SpatRast for one point
lon = -86.2621555581 #El Bramadero lon (from NicaAgua)
lat = 13.3816217871 #El Bramadero lat (from NicaAgua)
lonLat = data.frame(lon=lon,lat=lat)

# Set up precipitation data
location = terra::vect(lonLat, crs = "+proj=longlat +datum=WGS84")
precip = terra::extract(infile, location, method = 'bilinear') %>%
  subset(select = -ID) %>%
  t()
precip[precip < 0] <- 0 #replace any negative (errant) values with zeroes
precip = as.vector(precip)
```

Because we will be using the msdGraph function, a year value will need to be provided. We will create a dummy year that will be used for the sake of plotting. This can be any year, but it is easiest to just use the first year in the data set.

```{r}
# Set up dates (time) data for ONE year (365 days)
allTimes = terra::time(infile) %>%
  as.Date() %>% 
  data.frame()
timeFrame = as.Date(c(allTimes[1,1]:allTimes[365,1])) %>% 
  data.frame()
# Make a Dataframe for each year
chunkLength = 365
divYears = (split(precip,ceiling(seq_along(precip) / chunkLength)))
divYearsFrame = data.frame(divYears)

averagePrecip = c()
for (i in 1:nrow(divYearsFrame)) {
  daySum = sum(divYearsFrame[i,])
  averagePrecip = rbind(averagePrecip, daySum)
}
nyears = round(length(precip)/365)
averagePrecip = averagePrecip/nyears #(average over number of years)
```

Now, an averaged xts object can be created. This will display the data set's average rainfall over the dummy year.

```{r}
# Assemble the Timeseries
timeseriesFrame = cbind(timeFrame, averagePrecip)
colnames(timeseriesFrame) = c("Date", "Precipitation")
x = xts(timeseriesFrame$Precipitation, timeseriesFrame$Date) #This produces the "xts" screenshot
```

Now, perform the MSD calculations and plot the averaged graph!

```{r}
# Perform MSD calculations with the new xts
keyDatesTS = msdDates(time(x))
filterTS = apply(x, MARGIN = 2, FUN = msdFilter, window = 31, quantity = 2)
duration <- apply(filterTS, MARGIN = 2, FUN = msdStats, keyDatesTS, fcn="duration")
intensity <- apply(filterTS, MARGIN = 2, FUN = msdStats, keyDatesTS, fcn="intensity")
firstMaxValue <- apply(filterTS, MARGIN = 2, FUN = msdStats, keyDatesTS, fcn="firstMaxValue")
secondMaxValue <- apply(filterTS, MARGIN = 2, FUN = msdStats, keyDatesTS, fcn="secondMaxValue")
minValue <- apply(filterTS, MARGIN = 2, FUN = msdStats, keyDatesTS, fcn="min")
allStats = msdMain(x)
graph1981 = msdGraph(x, 1981)
plot(graph1981)
```

Note that the first 31 days and the final 31 days will not have values due to the behavior of the msdFilter function with a single "year" of data.
