---
title: "msdMain-with-raster"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{msdMain-with-raster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MSD)
```

In general, the MSD Package works best when the data from a specific point in a raster is pulled, converted into an xts object, and then analyzed. This vignette will show an alternative approach, when an entire raster is analyzed using the msdMain function, then a single specific point and its corresponding data is extracted to a dataframe.

```{r}
# Load the necessary packages and the included data, for the purpose of demonstration
library(MSD)
library(terra)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(xts)
data = system.file("data", "prcp_cropped.tif", package="MSD") #This loads the data included in the package, but you would attach your own
infile = terra::rast(data)
```

Now we have our raster data loaded in as "infile". Using the terra::app command, we can apply a function to each cell of the data. The msdMain function applies the msdDates, msdFilter, and msdStats function all in one, so it can be directly called once the data is read in.

```{r}
# Run the msdMain function across the entire raster
# msdMain(x, firstStartDate, firstEndDate, secondStartDate, secondEndDate, window, quantity, timeVector)
fSD = "05-01" #firstStartDate
fED = "08-31" #firstEndDate
sSD = "06-01" #secondStartDate
sED = "10-31" #secondEndDate
w = 31 #window
q = 2 #quantity
# WORK IN PROGRESS
#allStatsRaster <- terra::app(x = infile, msdMain, fSD, fED, sSD, sED, w, q)
```

