---
title: "multiple-years"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multiple-years}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MSD)
```

The basic function of the MSD package is to create a timeseries plot of smoothed rainfall data in order to highlight the important characteristics and determine if an MSD is present in a single year. With some manual work, this functionality can be expanded so that multiple years of rainfall data can be displayed side by side. This vignette will guide you through the process of taking a single spatial point of rainfall data over a series of years and plotting those side-by-side.

Begin by loading in your data, either in the form of a raster or an xtimeseries.

```{r}
library(terra)
library(tidyr)
library(lubridate)
library(stringr)
library(ggplot2)
library(xts)
data = system.file("data", "prcp_cropped.tif", package="MSD") #This loads the data included in the package, but you would attach your own

# Extract 1 spatial point from the raster data
infile = terra::rast(data)
lon = -86.2621555581 #Longitude of the spatial point we're interested in analyzing
lat = 13.3816217871 #Lattitude of the spatial point we're interested in analyzing
lonLat = data.frame(lon=lon,lat=lat)

# Set up precipitation data by extracting the data located at our longitude and lattitude
location = vect(lonLat, crs = "+proj=longlat +datum=WGS84")
precip = terra::extract(infile, location, method = 'bilinear') %>%
  subset(select = -ID) %>%
  t()
precip[precip < 0] <- 0 #replace any negative (errant) values with zeroes
precipFrame = data.frame(precip)

# Set up dates (time) data which will be used in creating our time series
timeFrame = terra::time(infile) %>%
  as.Date() %>%
  data.frame()
startDate = as.Date(as.character(timeFrame[1,1]))
endDate = as.Date(as.character(timeFrame[nrow(timeFrame),1]))
datesSequence = seq(from = startDate, to = endDate, by = 1)
timeseriesFrame = cbind(timeFrame, precipFrame)
colnames(timeseriesFrame) = c("Date", "Precipitation")

# Make the data into an xtimeseries that the package recognizes
x = xts(timeseriesFrame$Precipitation, timeseriesFrame$Date)
```

At this point, we taken a raster of data and extracted from it a series of precipitation values and a series of time for a single location. This data can now be used with the MSD package and its functions. Alternatively, as mentioned above, you could directly load in xtimeseries data. The workflow for that is much simpler than that of loading and preparing a raster.

```{r}
data("timeseries") #This loads the data included in the package, but you would attach your own
x = timeseries
```

The end result of either of these processes is a time series that has rainfall data over a set of dates. Here is the plot of the xtimeseries.

```{r fig.height=5, fig.width=5}
# Plot the xtimeseries so we have a visualization of our task
plot(x)
```

As you can see, this data set has data for 1981, 1982, 1983, 1984, and 1985. In general, the MSD package can analyze one of these years at a time. In order to analyze multiple years at once, we'll need to create loops, extract the data from each individual (single-year) graph, then create a new plot with all the data together.

```{r}
#combinedPlots = NULL
#firstYear = lubridate::year(startDate)
#lastYear = lubridate::year(endDate)
#for (i in firstYear:lastYear) {
  #onePlot = msdGraph(x, i)
  #plotData = ggplot_build(onePlot)$plot$data
  #plotColor = ggplot_build(onePlot)$plot$plot_env$color1
  #allPlot = data.frame(plotData, plotColor)
  #combinedPlots = rbind(combinedPlots, allPlot)
#}
#colnames(combinedPlots) = c("Date", "Precip", "Color")
#combinedPlotData = data.frame(as.Date(combinedPlots$Date), as.numeric(combinedPlots$Precip))
#colnames(combinedPlotData) = c("Date", "Precip")

```

Now let's use ggplot to create a new combined graph

```{r}
#finalPlot = ggplot(data = combinedPlotData, mapping = aes(x=Date,y=Precip))+
  #geom_line(data=combinedPlotData[1:nrow(combinedPlotData),])+
  #xlab("")+
  #ylab("Precipitation (mm/day)")+
  #theme_bw()+
  #annotate("rect", xmin=combinedPlotData[1,1], xmax=combinedPlotData[nrow(combinedPlotData),1], #ymin=0, ymax=(max(combinedPlotData$Precip, na.rm=TRUE)+2), fill="white", alpha=0.5)
#!! WIP: fill needs to be changed, also needs to have the lines and dots
```

Now we can view the MSD data for all the years of rainfall data we have. If you only want to see a certain range of data, simply replace startYear and endYear with the years you're interested in.

Let's view our combined plot!

```{r fig.height=5, fig.width=5}
plot(finalPlot)
```
